Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-08-03 document/ms.tex
Closes #32-- shorten Section 3.2

An archive of 1216 $B$, $V$ and/or $R$ photometric monitoring visits was assembled from published and unpublished data sources spanning 31 years in 22 observing seasons.  Published data targeting \name is composed of 29 $BVR$ visits from 1985$-$1986 \citep{vrba93}, 26 $BVR$ epochs from 1990$-$1991 \citep{bouvier93}, 284 visits in $V$-band (278 with $B$ and 268 with $R$) from 1992 August - 2004 October \citep{grankin08}, and 10 $BVR$ visits from 2013 \citep{donati14}.  ASAS3 \citep{pojmanski04} acquired 63 $V-$band measurements from 2002$-$2004.  Integral-OMC obtained 138 $V-$band measurements from 2006$-$2008 \citep{garzon12}.  The AAVSO archive \citep{kafka16} includes 385 $V$, 23 $B$ and 10 $R$ measurements from 2013$-$2016.  Unpublished data from the ASAS-SN survey \citep{shappee14}, were obtained from 2012 January -- 2016 March in 186 visits.  Finally, recent photometry from the ongoing Crimean Astrophysical Observatory (CrAO) ROTOR project \citep{grankin08} is presented here for the first time.  The CrAO data includes 43 $BVR$ visits from August 2015 - March 2016. Figure \ref{fig:PhotTime} displays all the $V-$band photometry over the interval 1985-2016.

@@

Figure \ref{fig:PhotPhase} shows all available $V-$band data grouped by the 22 observing seasons and phase-folded by period $P=3.375$ days obtained from multiterm Lomb-Scargle periodograms \citep{ivezic14}.  The general appearance of the phase-folded lightcurves does not change with perturbations to the period on the scale of 0.003 days.  The estimated $BVR$ magnitudes at the time of the spectral observations are determined from a regularized multiterm fit \citep{vanderplas15a} shown as the solid blue line in Figure \ref{fig:PhotPhase}.  Table \ref{tbl_estimated_V} lists the estimated $BVR$ photometry and the observing epoch for data from IGRINS, ESPaDOnS, 2MASS \citep{skrutskie06}, DoubleSpec, and TripleSpec.

@@

\citet[hereafter \iancze]{czekala15} developed a modular framework\footnote{The open source codebase and its full revision history is available at \url{https://github.com/iancze/Starfish}.  The experimental fork discussed in this paper is at \url{https://github.com/gully/Starfish}} to infer stellar properties from high resolution spectra.  The \iancze\ technique forward models ``emulated'' synthetic spectra from pre-computed model grids, taking into account the discretization error attributable to the coarsely sampled stellar intrinsic parameters.  We employed the \PHOENIX\ grid of pre-computed synthetic stellar spectra that span a wide range of wavelengths at high spectral resolution with sampling of 100 K in $\teff$ in our range of interest \citep{husser13}.  The modular framework was altered to accommodate starspot measurements in two ways. First, the single photospheric component was updated to includes a starspot spectrum. Second, the MCMC sampling strategy was altered to accommodate the additional free parameters added in the starspot model.

@@

The Metropolis-Hastings step sizes were tuned with several iterations of burn-in procedure and the final chains were visually checked for convergence.  Nine of the 35 available spectral orders failed to converge, possibly due to numerical artifacts arising from poor model fits.  For the other 26 spectral orders, the number of samples was generally much longer than the estimated integrated autocorrelation length.  Overlays of models with observed spectra show modest agreement, with the exception of conspicuous spectral line outliers.  No attempt was made to downweight spectral-line outliers as described in \iancze.

We computed the median value and 5$^{th}$ and $95^{th}$ percentiles of burned-in subsets of the MCMC samples described above.  Overall these 26 sets of point estimates for $\vT_{o}$, $\vP_o$ show relatively good agreement, with exceptions.  We have little constraint on $\logg$ and $\Z$ since full-spectrum fitting tends to wash out the weak signal contained in line wings for spectra with a finite level of random noise and systematic error.  Meanwhile, strong metal lines can bias estimates of $\Z$ at the expense of estimates for $\teff$.  We get consistent estimates for $v_z \sim 15.6 \pm 0.6$ km/s and $\vsini \sim 28.0 \pm 0.9$ km/s, with the exception of two spectral orders at $6920 <\lambda \;(\mathrm{nm})< 7057$ and $8473 < \lambda \;(\mathrm{nm}) < 8706$.  These two outlying spectral orders converged to unreasonably large values for $\vsini$ and $a_{\rm G}$, the amplitude of the Gaussian process covariance kernel.  One of these orders has an unusually small value for effective temperature, $\teff\sim3500$, comparable to the lower limit of the employed model grid.  Visual comparison of the observed spectrum for a range of fits for these orders shows that no single emulated spectrum could reasonably fit their spectrum because there are numerous large spectral outliers.  The more-well-behaved spectral orders show effective temperature point estimates in the range $\teff=4000\pm130$ K.  Figure \ref{fig:SingleTeffvsOrder} displays the $\teff$ point estimates with 5$^{th}$ and $95^{th}$ percentile error bars placed at the central wavelength of each of the 26 spectral orders.

\begin{figure*}
	\centering
	\includegraphics[width=0.8\textwidth]{figures/single_Teff_v_order}
	\caption{Effective temperature as derived from unique full spectrum fitting to each of 58 spectral orders in the optical through infrared portions of the spectrum and assuming a single component photosphere.  The top panel indicates the location and size of spectral chunks used to derive the values in the bottom panel.  The IGRINS orders also show typical atmospheric transmission (red lines) in the infrared.  Unused spectral chunks are lightened.  The effective temperature derived in the $K-$band is about 800 K lower than that derived in optical.}
	\label{fig:SingleTeffvsOrder}
\end{figure*}

@@

We performed full-spectrum fitting on 32 of the 54 available IGRINS spectral orders, again fitting unique stellar parameters $\vt = (\teff, \logg, \Z)$ for each spectral order $o$.  We used the same analysis procedure as described for the ESPaDOnS spectra, with one exception.  For the IGRINS $K-$band, we employed an expanded search range for the effective temperature, $\teff \in [3000, 4200]$ K, since the IGRINS $H-$ band demonstrated some saturation at the $\teff=3500$ K lower bound.  No effort was made to shift the wavelength scales of the IGRINS data and \PHOENIX\ model into the same system, so the absolute values of $v_z$ are arbitrary shifted.

We find a larger dispersion in the point estimates for the stellar parameters derived from the IGRINS data with single temperature fits.  The most conspicuous trend is in the derived effective temperature as a function of wavelength shown in Figure \ref{fig:SingleTeffvsOrder}.  Here we see dramatic result that the effective temperature peaks at a values of $\sim4200$K in the short wavelength end of $H-$band and saturates at $<3500$ K at the long wavelength end of $H-$band.  The $K-$band shows even lower derived effective temperatures of $\sim3300$ K.

\subsection{Interpretation of the single temperature fitting results}

The effective temperature derived at different wavelength intervals exhibit a $\Delta \teff \sim 800$ K spread.  No single temperature can describe all the spectral lines present in the high resolution optical and IR spectra.  This discrepancy is circumstantial evidence for the detection of spectral lines attributable to starspots.  This circumstantial evidence should not be surprising given the previous evidence that \name possesses a large coverage fraction of starspots \citep{grankin08,donati14}.

%We performed a further plausibility check to constrain the effect size of starspots.  We generated flux calibrated spectra at two temperatures, and forward-modeled them to resemble the LkCa4 IGRINS and ESPaDOnS spectra in all other ways (\emph{i.e.} $\logg, \Z, \vt_{\rm ext}$), including noise and calibration parameters.  We coadded the spectra in a mixture model:  $ \mathsf{M}_{mix} = c \cdot \mathsf{M}_A(T_\textrm{eff,A}) + (1-c) \cdot \mathsf{M}_B(T_\textrm{eff,B})$.  We chose a fill factor of starspots of 30\%.  We then re-ran our single-temperature fitting procedure on the two-temperature, synthetic, noised-up data to see what stellar parameters our full-spectrum fitting procedure would na\"{\i}vely derive.

@@

%In Figure XX we plot the ratio of black body flux density at temperatures 4100 K and 3300 K with a red line.  The blue line is the ratio of two smoothed \PHOENIX model spectra with 4100 K and 3300 K, solar metallicity, and $\logg=3.5$.  The flux density ratio is the flux per solid angle, so the delivered flux ratio for a finite sized starspot will depend on the (flux-weighted) solid angle of the starspot relative to the (flux-weighted) solid angle of the stellar disk, vis-a-vis Equations \ref{eqn:Mix_M} and \ref{eqn:fill_factor}.  
%\todo{Put in the flux density ratio plot.}
%\comm{Should we change the plot numbers from 4100 and 3300 to 4200 and 3000?  -These would end up being closer to the best fit values.}

In summary, the visible portion of the spectrum will be dominated by the warm patches, but the contrast with the starspots will be higher.  The infrared spectrum will have a lower overall spot contast, but will have higher relative contribution attributable to the cool patches.  This observation has motivated precision radial velocity (RV) planet searches in the infrared, where the modulation of RV signals is lessened \citep{prato08}.
%\comm{This blurb should probably go in the introduction.}

@@

Full-spectrum fitting was performed for 43 of the 54 available IGRINS spectral orders as described in Section \ref{sec:methods} and Appendix \ref{methods-details}.  The MCMC samples had 5000 steps with 40 walkers.  The stellar parameter ranges were $\logg \in [3.5, 4.0]$ and $\Z \in [-0.5, 0.5]$; the effective temperature range for $H-$ band was $\teffa, \teffb \in [2700, 4500]$, and was expanded to $\teffa, \teffb \in [2700, 4500]$ for $K-$band.  The MCMC chains all appeared to converge after about 1500 steps.

Marginalized point estimates were computed for all 14 stellar and nuisance parameters by selecting the final 200 $\times$ 40 walkers = 8000 samples, and computing the median, $5^{th}$ and $95^{th}$ percentiles of the marginal distributions.  The fit quality was assessed by examining the consistency of the point estimates across the spectral orders.  The distribution of $v_z$ and $\vsini$ were informative for finding extremely poor fits.  The range for all 43 orders was $v_z = 12.4 \pm 2.6$ and $\vsini = 28.8 \pm 2.0$.  Spectral orders $o=91$ and $94$ had exceptionally large uncertainties on their values for $v_z$ and $\vsini$, which makes sense since these two orders are badly affected by telluric absorption artifacts.

As noted previously, the full-spectrum fitting method offers little constraining power on $\logg$, and $\Z$ can be biased in the face of systematic spectral line outliers.  Estimates of $\logg$ appeared the high end of our search range, $3.5 <\logg <4.0$, yielding a median value of the point estimates of 3.8 and a standard deviation of 0.1, consistent with previously published values from \citet{donati14}.  The estimates for $\Z$ yield a slight trend with wavelength-- the longest wavelength orders tend to prefer higher metallicity, while the shorter wavelength values prefer slightly sub-solar metallicity.  Orders 79, 81, 90, and to a lesser extent 78 and 76 showed artificially tight constraints on $\Z$, biased to large values.  Similarly, orders 86, 101, 102, 104, 106, all had $\Z$ estimates biased to extremely low values.   Visual inspection of these orders indicates that their spectrum is dominated by spectral line outliers--deep metal lines that cannot be described by a coarse metallicity estimate alone.  These spectral line outlier orders also tend to bias estimates of effective temperature.  These metal lines can also be biased by Zeeman broadening, which would explain the heightened effect size with wavelength.

@@

{\it Facilities:} \facility{Smith (IGRINS)}, \facility{AAVSO}, \facility{CFHT (ESPaDOnS)}, \facility{INTEGRAL (OMC)}, \facility{ASAS}, \facility{CrAO:1.25m}, \facility{ARC (TripleSpec)}, \facility{Hale (DBSP)}
{\it Software: } TBD
	\project{pandas} \citep{mckinney10}
    \project{emcee} \citep{foreman13},
	\project{matplotlib} \citep{hunter07},
	\project{numpy} \citep{vanderwalt11},
	\project{scipy} \citep{jones01},
	\project{ipython} \citep{perez07},
	\project{gatspy} \citep{JakeVanderplas2015},
	\project{starfish} \citep{czekala15}
%\software{%
%	\project{pandas} \citep{mckinney10}
%    \project{emcee} \citep{foreman13},
%	\project{matplotlib} \citep{hunter07},
%	\project{numpy} \citep{vanderwalt11},
%	\project{scipy} \citep{jones01},
%	\project{ipython} \citep{perez07},
%	\project{gatspy} \citep{JakeVanderplas2015},
%	\project{starfish} \citep{czekala15}}.

>>>>

An archive of 1216 $B$, $V$ and/or $R$ photometric monitoring visits was assembled from published and unpublished data sources spanning 31 years in 22 observing seasons.  Published data targeting \name is composed of 29 $BVR$ visits from 1985$-$1986 \citep{vrba93}, 26 $BVR$ epochs from 1990$-$1991 \citep{bouvier93}, 284 visits in $V$-band (278 with $B$ and 268 with $R$) from 1992 August - 2004 October \citep{grankin08}, and 10 $BVR$ visits from 2013 \citep{donati14}.  ASAS3 \citep{pojmanski04} acquired 63 $V-$band measurements from 2002$-$2004.  Integral-OMC obtained 138 $V-$band measurements from 2006$-$2008 \citep{garzon12}.  The AAVSO archive \citep{kafka16} includes 385 $V$, 23 $B$ and 10 $R$ measurements from 2013$-$2016.  Unpublished data from the ASAS-SN survey \citep{shappee14}, were obtained from 2012 January -- 2016 March in 186 visits.  Finally, recent photometry from the ongoing Crimean Astrophysical Observatory (CrAO) ROTOR project \citep{grankin08} is presented here for the first time.  The CrAO data includes 43 $BVR$ visits from August 2015 - March 2016. Figure \ref{fig:PhotTime} displays all the $V-$band photometry over the interval 1985-2016.  HJD, BJD, and JD are simply refered to as JD, which is accurate to the precision of available data.

@@

Figure \ref{fig:PhotPhase} shows all available $V-$band data grouped by the 22 observing seasons and phase-folded by period $P=3.375$ days obtained from multiterm Lomb-Scargle periodograms \citep{ivezic14}.  The general appearance of the phase-folded lightcurves does not change with perturbations to the period on the scale of 0.003 days.  The estimated $BVR$ magnitudes at the time of the spectral observations are determined from regularized multiterm fits \citep{vanderplas15a} shown as the solid blue lines in Figure \ref{fig:PhotPhase}.  Table \ref{tbl_estimated_V} lists the estimated $BVR$ photometry and the observing epoch for data from IGRINS, ESPaDOnS, 2MASS \citep{skrutskie06}, DoubleSpec, and TripleSpec.

@@

\citet[hereafter \iancze]{czekala15} developed a modular framework\footnote{The open source codebase and its full revision history is available at \url{https://github.com/iancze/Starfish}.  The experimental fork discussed in this paper is at \url{https://github.com/gully/Starfish}} to infer stellar properties from high resolution spectra.  The \iancze\ technique forward models ``emulated'' synthetic spectra from pre-computed model grids, taking into account the discretization error attributable to the coarsely sampled stellar intrinsic parameters.  We employed the \PHOENIX\ grid of pre-computed synthetic stellar spectra that span a wide range of wavelengths at high spectral resolution with sampling of 100 K in $\teff$ in our range of interest \citep{husser13}.  The modular framework was altered to accommodate starspot measurements in two ways. First, the single photospheric component was updated to include a starspot spectrum. Second, the MCMC sampling strategy was altered to accommodate the additional free parameters added in the starspot model.

@@

The Metropolis-Hastings step sizes were tuned with several iterations of burn-in procedure and the final chains were visually checked for convergence.  Nine of the 35 available spectral orders failed to converge, due to numerical artifacts arising from poor model fits.  For the other 26 spectral orders, the number of samples was generally much longer than the estimated integrated autocorrelation length.  Overlays of models with observed spectra show modest agreement, with the exception of conspicuous spectral line outliers.  No attempt was made to downweight spectral-line outliers as described in \iancze.

We computed the median value and 5$^{th}$ and $95^{th}$ percentiles of burned-in subsets of the MCMC samples described above.  Overall the 26 sets of point estimates for stellar parameters show relatively good agreement, with exceptions.  We have little constraint on $\logg$ and $\Z$ since full-spectrum fitting tends to wash out the weak signal contained in line wings for spectra with a finite level of random noise and systematic error.  Meanwhile, strong metal lines can bias estimates of $\Z$ at the expense of estimates for $\teff$.  Estimates for $v_z \sim 15.6 \pm 0.6$ km/s and $\vsini \sim 28.0 \pm 0.9$ km/s are consistent across all orders with the exception of two spectral orders at $6920 <\lambda \;(\mathrm{nm})< 7057$ and $8473 < \lambda \;(\mathrm{nm}) < 8706$ due to numerous large spectral outliers.  The more-well-behaved spectral orders show effective temperature point estimates in the range $\teff=4000\pm130$ K.  Figure \ref{fig:SingleTeffvsOrder} displays the $\teff$ point estimates with 5$^{th}$ and $95^{th}$ percentile error bars placed at the central wavelength of each spectral order.

\begin{figure*}
	\centering
	\includegraphics[width=0.65\textwidth]{figures/single_Teff_v_order}
	\caption{Effective temperature as derived from unique full spectrum fitting to each of 58 spectral orders in the optical through infrared portions of the spectrum and assuming a single component photosphere.  The top panel indicates the location and size of spectral chunks used to derive the values in the bottom panel.  The IGRINS orders also show typical atmospheric transmission (red lines) in the infrared.  Unused spectral chunks are lightened.  The effective temperature derived in the $K-$band is about 800 K lower than that derived in optical.}
	\label{fig:SingleTeffvsOrder}
\end{figure*}

@@

We performed full-spectrum fitting on 32 of the 54 available IGRINS spectral orders, again fitting unique stellar parameters $\vt = (\teff, \logg, \Z)$ for each spectral order $o$.  We used the same analysis procedure as described for the ESPaDOnS spectra, with one exception.  For the IGRINS $K-$band, we employed an expanded search range for the effective temperature, $\teff \in [3000, 4200]$ K, since the IGRINS $H-$ band demonstrated some saturation at the $\teff=3500$ K lower bound.

We find a larger dispersion in the point estimates for the stellar parameters derived from the IGRINS data than those derived in optical.  The most conspicuous trend is in the derived effective temperature as a function of wavelength shown in Figure \ref{fig:SingleTeffvsOrder}.  The effective temperature peaks at a values of $\sim4200$K in the short wavelength end of $H-$band and saturates at $<3500$ K at the long wavelength end of $H-$band.  The $K-$band shows even lower derived effective temperatures of $\sim3300$ K, a $Delta \teff \sim 700$ K spread from optical.  No single temperature can describe all the spectral lines present in the high resolution optical and IR spectra.  This discrepancy is circumstantial evidence for the detection of spectral features attributable to starspots.

@@

In summary, the visible portion of the spectrum will be dominated by the warm patches, but the contrast with the starspots will be higher.  The infrared spectrum will have a lower overall spot contast, but will have more net starspot flux than optical.  This observation has motivated precision radial velocity (RV) planet searches in the infrared, where the modulation of RV signals is lessened \citep{prato08}.

@@

Full-spectrum fitting was performed for 43 of the 54 available IGRINS spectral orders as described in Section \ref{sec:methods} and Appendix \ref{methods-details}.  The MCMC samples had 5000 steps with 40 walkers.  The stellar parameter ranges were $\logg \in [3.5, 4.0]$ and $\Z \in [-0.5, 0.5]$; the effective temperature range was $\teffa, \teffb \in [2700, 4500]$.  $H-$band fits had normal distribution priors in place: solar metallicity to $\pm0.05$ dex, $logg=3.8\pm0.1$, and $\vsini=29\pm5$.  The MCMC chains all appeared to converge after about 1500 steps.

Marginalized distributions were computed for all 14 stellar and nuisance parameters by selecting the final 200 $\times$ 40 walkers = 8000 samples, and point estimates were obtained by computing the median, $5^{th}$ and $95^{th}$ percentiles of the marginal distributions.  The fit quality was first assessed by examining the consistency of the point estimates across the spectral orders.  The distribution of $v_z$ and $\vsini$ were informative for finding extremely poor fits.  The range for all 43 orders was $v_z = 12.4 \pm 2.6$ and $\vsini = 28.8 \pm 2.0$.  Two orders with poor telluric correction ($o=91$ and $94$) had exceptionally large uncertainties on their values for $v_z$ and $\vsini$.

Some orders indicates that their spectrum is dominated by spectral line outliers--deep metal lines that cannot be described by a coarse metallicity estimate alone.  These spectral line outlier orders also tend to bias estimates of effective temperature.  These metal lines can also be biased by Zeeman broadening, which would explain the heightened effect size with wavelength.

@@

{\it Facilities:} \facility{Smith (IGRINS)}, \facility{AAVSO}, \facility{CFHT (ESPaDOnS)}, \facility{INTEGRAL (OMC)}, \facility{ASAS}, \facility{CrAO:1.25m}, \facility{ARC (TripleSpec)}, \facility{Hale (DBSP)}
{\it Software: } TBD
	\project{pandas} \citep{mckinney10}
    \project{emcee} \citep{foreman13},
	\project{matplotlib} \citep{hunter07},
	\project{numpy} \citep{vanderwalt11},
	\project{scipy} \citep{jones01},
	\project{ipython} \citep{perez07},
	\project{gatspy} \citep{JakeVanderplas2015},
	\project{starfish} \citep{czekala15}
	\project{seaborn} \citep{waskom14}
%\software{%
%	\project{pandas} \citep{mckinney10}
%    \project{emcee} \citep{foreman13},
%	\project{matplotlib} \citep{hunter07},
%	\project{numpy} \citep{vanderwalt11},
%	\project{scipy} \citep{jones01},
%	\project{ipython} \citep{perez07},
%	\project{gatspy} \citep{JakeVanderplas2015},
%	\project{starfish} \citep{czekala15}}.
